---
title: "Mini “exploratory report” on juvenile snowshoe hares in Bonanza Creek Experimental Forest (Long Term Ecological Research site)"
subtitle: "Homework 4 - Task 2"
author: "Carmen Galaz-García"
date: "11/15/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)  # include packages 
library(lubridate)  #  handle dates
library(janitor)    # clean names (not used)
library(here)       # local paths to files
library(ggbeeswarm) # make beeswarm graphs
library(kableExtra)  # create nice tables
library(effsize)    # includes cohen's d

library(broom)
library(ggpubr)

hares<-read.csv(here("data","bonanza_hares.csv"))  # read data

# - selecting data for juvenile hares (used throughout)
juveniles <- hares %>% filter(age=="j")
```

## Introduction

## Data and analysis
Size measurements, sex and age of snowshoe hares were collected and made available by Dr. Knut Kielland and colleagues at the Bonanza Creek Experimental Forest Long Term Ecological Research (LTER) site located approximately 20 km southwest of Fairbanks, Alaska. The data contains observations of 3197 snowshoe hares obtained by capture-recapture studies conducted yearly from 1999 to 2012 in three sampling sites: Bonanza Riparian, Bonanza Mature and  Bonanza Black Spruce. 

The focus of this report is on observations of juvenile snowshoe hares.

-*Section 1) Annual juvenile hare trap counts*   
Here we explore the counts by year of juvenile snowshoe hares and include a table with summary statistics. 

-*Section 2) Juvenile hare weights by sex and sample site*
********

-*Section 3) Weight comparison by sex of juvenile hares*   
In this section we explore the weights of male and female juvenile hares in each sampling site. Differences between male and female hares are explored by a two-sample t-test using a significance level of 0.05 and by calculating Cohen's *d* effect size. 

-*Section 4) Relationship between juvenile weight & hind foot length*   
Finally we explore a possible relation between hind foot length and weight of juvenile hares by conducting a simple linear regression model on these variables and the fitness of this model is discussed. 

All analyses in R version 4.0.2 using RStudio Version 1.3.1093.


## Exploratory findings

### Section 1) Annual juvenile hare trap counts

Yearly counts of juvenile snowshoe hares are analyzed. 
In total 378 observations of juvenile snowshoe hares were recorded during the study.
Sample site is not included as a variable here, though it would be it would be interesting to see if juvenile hare counts are changing equally in all sample sites. 

```{r, echo=FALSE}
# --- Count juveniles per year
per_year_juveniles <- hares %>%    # copy data
  filter(age == "j") %>%       # select juvenile hares
  mutate(Year = year(mdy(date))) %>%     # create new column with year
  count(Year)

# --- Create graph
ggplot( data= per_year_juveniles, aes(x =Year, y =n)) +   # select data and variables
  geom_bar(stat="identity", fill="snow4") +    # do a bar plot
  scale_x_continuous(breaks=2*(994:1006)) +   # adjust ticks
  labs(#title="Juvenile snowshoe hares observations recorded yearly",
       y= "Number of hares (n)") +
  theme_light()  # update theme


```

**Figure 1.** Total juvenile snowshoe hare observations (n) during each year of the study (1999-2012). 
Data: Kielland et al. (2017)

No observations of juvenile snowshoe hares were made during 2002 and 2009, while the highest count was registered in 1999 (the first year of the study) with 
`r max(per_year_juveniles$n)` observations. 
On average `r mean(per_year_juveniles$n)` 
juvenile hare trappings were made from 1999 to 2012 and the median was `r median(per_year_juveniles$n)`. 
In figure one we can see a  diminishing trend in the number of juvenile hare trappings. The two exceptions are 2005, 2008 and 2011. 
In 2005 juvenile hare counts were 
`r per_year_juveniles[6,2]/per_year_juveniles[5,2]` times higher than in 2004, and in 2008 counts were
`r per_year_juveniles[9,2]/per_year_juveniles[8,2]` times higher than in 2007. 
In 2011 the juvenile snowshoe hare count was 
`r per_year_juveniles[11,2]/per_year_juveniles[10,2]` times higher than in 2010.

While these are absolute counts, these numbers are also impacted by effort. Moving forward I would suggest to conduct the trappings on the same dates every year and space these times across different seasons. For example the first two weeks of each month every year. I imagine juveniles are mostly present during Spring, but with climate change it is possible Spring weather does not come at the same dates in 2012 as in 1998, so there might be a risk of undercounting if the counts are done during a fixed period of time every year. I would also suggest to maintain the number and location of traps constant over the years. 


### Section 2) Juvenile hare weights by sex and sample site

```{r, message=FALSE, echo=FALSE}
# - select data
summary_juveniles <- juveniles %>% 
  group_by(grid, sex) %>% 
  summarize(mean = mean(weight, na.rm=TRUE),
            sd = sd(weight, na.rm=TRUE),
            #median = median(weight, na.rm=TRUE),
            n = n()
            )

```

Weights (measured in grams) of male and female juvenile snowshoe hares are compared across the three sampling sites: Bonanza Black Spruce, Bonanza Mature and Bonanza Riparian. 
From the 378 observations of juvenile snowshoe hares recorded during the study, 7 of them did not have the hare's weight recorded (3 females and 4 of unidentified sex). 


In all three sites the average weight for male juvenile hares is greater than that of female juvenile hares. 
Juvenile hares were biggest at the Bonanza Black Spruce sampling site, with weight of male hares being 
`r round(summary_juveniles$mean[2])` $\pm$ 
`r round(summary_juveniles$sd[2])` grams
(n= `r summary_juveniles$n[2]`; mean $\pm$ 1 standard deviation) and weight of female hares being
`r round(summary_juveniles$mean[1])` $\pm$ 
`r round(summary_juveniles$sd[1])` grams
(n= `r summary_juveniles$n[1]`). 
In Bonanza Mature and Bonanza Riparian the average and standard deviations of weights were similar for both female and male juvenile hares. 

**Table 1.** Descriptive statistics (mean, standard deviation and sample size) at each sample site for female and male juvenile snowshoe hares. Data: Kielland et al. (2017)

```{r message=FALSE, warning=FALSE, echo= FALSE}

# --- Create table with summar statistics 

weight_summary_table <- summary_juveniles %>% 
  mutate(
       grid = case_when( # -- update names of sample sites
         grid == "bonbs" ~ "Bonanza Black Spruce",
         grid == "bonrip" ~ "Bonanza Riparian", 
         grid == "bonmat" ~ "Bonanza Mature"
       ),
       sex = case_when(
         sex == "f" ~ "female",
         sex == "m" ~"male"
       )) %>% 
  kable( col.names = c("Sample site", 
                        "Sex", 
                        "Mean weight (g)",
                        "Standard deviation (g)",
                        #"Median weight (g)",
                        "Sample size (n)" ) 
                         ) %>% 
   kable_styling(full_width = FALSE)
weight_summary_table
```

Distribution of weights for female and male juvenile hares appear to be more normal at the Bonanza Riparian site and rather uniform at the other two sites. 
Note that the Bonanza Riparian site has almost four times as many observations than the other two sites for female hares. 
For male hares Bonanza Riparian site has about twice as many observations than Bonanza Mature and almost 10 times more than Bonanza Black Spruce. 
In all three sites and for both sexes the eight mean and median are similar. 

```{r, message=FALSE, echo=FALSE, warning=FALSE}
# ---  Create graph of juvenile weight divided by sex and grid

# -- individual figure titles
grid_names <- c(bonbs="Bonanza Black Spruce", 
                bonrip="Bonanza Riparian", 
                bonmat="Bonanza Mature")

# - make beeswax plot with added box-plot on top
ggplot(data = juveniles, aes(x = sex, y = weight), ) +
  geom_beeswarm(show.legend=FALSE,
                aes(color=sex),
                size = 4,    # point size
                alpha = 0.6, # transparency
                pch = 20,    # point shape
                cex=2.5      # scatters points
                ) +
  geom_boxplot(fill = NA, width = 0.2, outlier.color = NA) +
  stat_summary(fun=mean, 
               geom="point", 
               shape=20, 
               size=3, 
               color="black", 
               fill="black")+
  facet_wrap(~grid,   # make one graph per grid element
             labeller = labeller( grid = grid_names)) +   # update individual titles     
  labs(y="weight (g)") +
  scale_x_discrete(labels = c("female", "male", "NA")) +   # update x-axis labels 
  theme_light()  # update theme

```

**Figure 2.** Weight (g) of juvenile snowshoe hares recorded in each sampling site, divided by sex. 
Red (female), blue (male) and gray (unidentified) points indicate individual observations of weight (g) of a juvenile snowshoe hare.
Box endpoints indicate the 25^th^ and 75^th^ percentile values, 
the black line and black point within the box indicate the median and mean value for each species, respectively.
Data: Kielland et al. (2017)



### iii) Juvenile weight comparison (male & female snowshoe hares).
Further compare mean weights for **juvenile male and female snowshoe hares** included in this study. In this section, you should include:
● A finalized table containing the mean, standard deviation, and sample size for male and female juvenile snowshoe hares, with a table caption (remember, this is placed *above* the table in your report)
● A 2 - 3 sentence comprehensive statement of the means comparison, including at least:
○ The actual difference in means
○ Effect size (Cohen's *d*)
○ Outcome of a two-sample t-test

```{r}
# --- checking how many hares don't have weight registered
#View(filter(juveniles,is.na(weight)))    # seven juvenile hares with weight==NA

# --- separate data of female and male juvenile hares with recorded weight
female_male <-  juveniles %>%   # copy data
  filter(!is.na(weight)) %>%           # select those with recorded weight
  filter(!is.na(sex))                  # select those with recorded sex

# --- create summary data for weight of female and male hares
female_male_summary <- female_male %>%   # copy data
  group_by(sex) %>% 
  summarize(mean_weight = mean(weight, na.rm=TRUE),
            sd_weight = sd(weight, na.rm=TRUE),
            median_weight = median(weight, na.rm=TRUE),
            sample_size = n())

# --- Update tags for table
female_male_summary[1,1] <- "female"  
female_male_summary[2,1] <- "male"  

# --- create table
female_male_summary %>% kable( col.names = c("Sex", 
                                       "Mean weight (g)", 
                                       "Standard deviation (g)",
                                       "Median weight (g)",
                                       "Sample size (n)" ) 
                         ) %>% 
   kable_styling(full_width = FALSE)
```


```{r, include=FALSE}
# --- Exploring data to ensure 2-sample t-test is adequate

# Histograms: female hares normal, male hares too though slightly bimodal
ggplot(data = female_male, aes(x = weight)) +
  geom_histogram(bins = 25) +
  facet_wrap(~sex)

# QQ Plots: both graphs close to a line
ggplot(data= female_male, aes(sample = weight)) +
  geom_qq() +
  facet_wrap(~sex)

```



```{r}
# --- Two-sample t-test & Cohen's d ----

# --- Convert data to vector form
juvenile_female <- female_male %>% 
  filter(sex == "f") %>% 
  pull(weight)

juvenile_male <- female_male %>% 
  filter(sex == "m") %>% 
  pull(weight)

# --- Two-sample t-test
ttest_juveniles <- t.test(juvenile_female, juvenile_male)
ttest_juveniles

# --- Cohen's d
cohen.d(juvenile_female, juvenile_male)
```

### iv) 
Relationship between juvenile weight & hind foot length. ​Explore the relationship between juvenile snowshoe hare hind foot length (variable `hindft`, recorded in ​millimeters)​ , and weight. Include all juvenile hares in the study. Create a final visualization of juvenile hare hind foot length versus weight. Include a figure caption.

```{r}
# --- Relationship between juvenile hind foot length and weight

# --- select data with recorded weight and recorded hindft
complete_juveniles <- juveniles %>% 
  filter(is.na(weight)==FALSE & is.na(hindft)==FALSE)
# View(filter(juveniles, is.na(weight) | is.na(hindft))) # checking the leftout data

# --- Graph
ggplot(data= complete_juveniles, aes(y=weight, x=hindft))+
  #geom_point( aes(col=sex), alpha = 0.4) +
  geom_point(  alpha = 0.4) +
  labs( y="Weight (g)", 
        x="Hind foot length (mm)")+
  theme_light()  # update theme


```

Upon seeing your graph, your “boss” tells you to try out​ linear regression to describe how juvenile snowshoe hare weight changes with respect to hind foot length. Try it out (the ​only​ variables you are expected to include here are hind foot length and weight, using data for all juvenile hares in the data, e.g. you are not expected to explore impacts of sex, site, etc.), and check the model diagnostics.
In 2 - 3 sentences below your figure, describe the outcome of your linear 2​
regressionexploration(e.g.slopeinterpretation,R​ value,andPearson's​r correlation) and touch briefly on which assumption(s) of linear regression
may be a concern, and additional thoughts you have about how well (or not well...) the linear model describes the relationship. What worries you?


```{r}
# --- Linear model for weight(hindft)
juveniles_lm <- lm(weight ~ hindft, data=complete_juveniles)
summary(juveniles_lm)
```

Multiple R-squared is 0.2988, which means that 29.88% of variance in  weight is explained by hind foot length. 

```{r}
# --- Accessing data in linear model

# model outputs converted to data frame
tidy_juveniles_lm <- broom::tidy(juveniles_lm) 

```

### Exploring model assumptions

```{r}
# --- Checking model assumptions
plot(juveniles_lm)

```

```{r}
# --- Visualize the model
ggplot( data=complete_juveniles, aes(x=hindft, y=weight)) +
  geom_point(alpha=0.4)+
  geom_smooth( method = "lm",
               color= "red",
               size =0.5, 
               fill="gray10",
               alpha=0.5)+
  theme_light()
```

```{r}
# --- Finding Pearson's r for correlation
juveniles_cor <- cor.test(complete_juveniles$hindft, complete_juveniles$weight)
juveniles_cor
```

